---
title: "Climate data Summary"
output: html_document
date: "`r format(Sys.time(), '%Y-%m-%d')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Data Description
- Author:  JuChien, Yao
- Date first created: 20251104
- Background: This dataset contains daily (or 3-daily) PM2.5 and AQI data for 4 different spots in Manhattan.
- Unit of observation: Daily, 3-daily
- Sample period: January 2024 - July 2025 (some up to March 2025)
- Known issues: None reported
- Definition for each variable: 
    - Date: year, month, date (a datetime object)
    - region: a string indicating the stand of the observed data (four spots included in Manhattan)
    - PM2.5: Average PM2.5 concentration of the day
    - AQI: Average AQI of the day
    - Daily Ridership: The total ridership for the day 

```{r}
# ---- Draft R translation (please review) ----
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
```

## Climate data (Air quality data)

```{r}
base_path = "C:/114-1_Autumn/Data_science_and_social_inquiry/clone/fa-25-econ-5166-group-3/data/processed"
input_data_file = file.path(base_path, "processed_pm2.5_AQI.csv")
```

```{r}
df = read.csv(input_data_file) %>% select(-X)
```

### Sample

```{r}
head(df)
```

### Summary Stats

```{r}
df$region %>% unique()
```

```{r}
df$PM2.5 %>%
  summary()
```

```{r}
df$AQI %>%
  summary()
```
