---
title: "hvfhv-dataset-summary-stats"
output: html_document
date: "`r format(Sys.time(), '%Y-%m-%d')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Data Description
- Author:  Yen-Chen Chou
- Date first created: 20251103
- Background: This dataset contains information of each ride of high-volume for-hire vehicles (HVFHVs) in New York City - Unit of observation: ride
- Sample period: January 2024 - August 2025
- Known issues: None reported
- Definition for each variable: 
    - hvfhs_license_num: The TLC license number of the HVFHS base or business.
      • HV0002: Juno
      • HV0003: Uber
      • HV0004: Via
      • HV0005: Lyft
    - request_datetime: Date/time when passenger requested to be picked up.
    - on_scene_datetime: Date/time when driver arrived at the pick-up location.
    - pickup_datetime : The date and time of the trip pick-up.
    - dropoff_datetime: The date and time of the trip drop-off.
    - PULocationID: TLC Taxi Zone in which the trip began.
    - DOLocationID: TLC Taxi Zone in which the trip ended.
    - trip_miles: Total miles for passenger trip.
    - trip_time: Total time in seconds for passenger trip.
    - base_passenger_fare: Base passenger fare before tolls, tips, taxes, and fees.
    - tolls: Total amount of all tolls paid in trip.
    - bcf: Total amount collected in trip for Black Car Fund.
    - sales_tax: Total amount collected in trip for NYS sales tax.
    - congestion_surcharge: Total amount collected in trip for NYS congestion surcharge.
    - airport_fee: $2.50 for both drop off and pick up at LaGuardia, Newark, and John F.Kennedy airports.
    - tips Total amount of tips received from passenger.
    - driver_pay: Total driver pay (not including tolls or tips and net of commission, surcharges, or taxes).
    - cbd_congestion_fee: Per-trip charge for MTA's Congestion Relief Zone starting Jan. 5, 2025.
  
```{r}
# ---- Draft R translation (please review) ----
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
```

## MTA Daily Dataset

```{r}
base_path = "~/Documents/DataScience/fa-25-econ-5166-group-3/data/processed"
input_data_file = file.path(base_path, "hvfhv_CBD_final.parquet")
```

```{r}
df <- read_parquet("/Users/alex/Desktop/FHVHV/hvfhv_CBD_final.parquet")
```

### Sample

```{r}
head(df)
```