---
title: "mta-hourly-dataset-summary-stats"
output: html_document
date: "`r format(Sys.time(), '%Y-%m-%d')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Data Description
- Author:  ChengYu, Lee
- Date first created: 20251103
- Background: This dataset contains hourly total ridership estimates for the MTA Subway, aggregated from minute-level data.
- Unit of observation: Hourly ridership estimates
- Sample period: July 2024 - June 2025
- Known issues: None reported
- Definition for each variable: 
    - Year: The year of the observation
    - Month: The month of the observation
    - Day of Week: The day of the week for the observation
    - Hour: The hour of the observation
    - Hourly Ridership: The total ridership for the hour

```{r}
# ---- Draft R translation (please review) ----
library(dplyr)
library(ggplot2)
library(readr)
```

## MTA Daily Dataset

```{r}
base_path = "~/Documents/DataScience/fa-25-econ-5166-group-3/data/processed"
input_data_file = file.path(base_path, "mta_hourly.csv")
```

```{r}
df = read_csv(input_data_file)
```

### Sample

```{r}
head(df)
```

### Summary Stats

```{r}
df$hour_sum %>%
    summary()
```

