---
title: "Process MTA data"
output:
  html_document:
    toc: true
    number_sections: false
    df_print: paged
---

```{r setup, include=FALSE}

```

# README

-   **Author**: `ChengYu, Lee`
-   **Created At**: `2025-09`
-   **Last Modified At**: `2025-11-03`

------------------------------------------------------------------------

## What does this file do?

-   Combine the raw data of MTA Subway Origin-Destination Ridership Estimate from 2024 and 2025 into a single dataset.
-   Filtering 2024 data to only include records from July onwards.
-   Save the combined dataset as a Parquet file for efficient storage and future analysis.
-   Generate hourly and daily total ridership estimates by aggregating the combined data, and save these summaries as CSV files.

------------------------------------------------------------------------

## What does this file take?

-   **Source Data Sets**:

    1.  `MTA_Subway_Origin-Destination_Ridership_Estimate__2024_20250928.csv`

    -   Description: `MTA Subway Origin-Destination Ridership Estimate data for the year 2024.`

    2.  `MTA_Subway_Origin-Destination_Ridership_Estimate__Beginning_2025.csv`

    -   Description: `MTA Subway Origin-Destination Ridership Estimate data for the year 2025.`

------------------------------------------------------------------------

## What does this file output?

-   `data.parquet` (intermediate file, not included in the repository)
    -   Description: `Combined MTA Subway Origin-Destination Ridership Estimate data for 2024 and 2025, filtered for July onwards in 2024.`
-   `mta_hourly.csv`
    -   Description: `Hourly total ridership estimates for the MTA Subway, aggregated from origin-destination pairs.`
-   `mta_daily.csv`
    -   Description: `Daily total ridership estimates for the MTA Subway, aggregated from hourly data.`

```{r}
library(data.table)
library(dplyr)
library(arrow)
library(ggplot2)
library(lubridate)
library(here)


base_path = "~/Documents/DataScience/fa-25-econ-5166-group-3"
```

```{r load_and_bind_data}
df_24 <- fread("~/Downloads/MTA_Subway_Origin-Destination_Ridership_Estimate__2024_20250928.csv")
df_25 <- fread("~/Downloads/MTA_Subway_Origin-Destination_Ridership_Estimate__Beginning_2025.csv")
# Note: As dataset of 2025 only cover first 6 months, we don't need to filter anymore.

df <- rbindlist(list(df_24, df_25), use.names = TRUE, fill = TRUE)

write_parquet(df, "~/Documents/DataScience/data.parquet", compression = "snappy")
```

```{r get_daily_total}
stations <- read.csv(paste0(base_path, "/data/raw/mta-cbd-stations.csv"))
stations <- as_arrow_table(stations) %>% 
    select(Name, in_CBD) %>%
    mutate(Name = cast(Name, large_utf8()))

ds_new <- open_dataset("~/Documents/DataScience/data.parquet", format = "parquet")
ds_new <- ds_new %>%
    left_join(
        stations %>% 
            rename(
                `Origin Station Complex Name` = Name,
                start_CBD = in_CBD
            ),
        by = "Origin Station Complex Name"
    ) %>%
    left_join(
        stations %>% 
            rename(
                `Destination Station Complex Name` = Name,
                end_CBD = in_CBD
            ),
        by = "Destination Station Complex Name"
    ) %>%
    select(
        -c(
            `Origin Station Complex ID`,
            `Origin Station Complex Name`,
            `Destination Station Complex ID`,
            `Destination Station Complex Name`,
            `Origin Latitude`, `Origin Longitude`,
            `Destination Latitude`, `Destination Longitude`,
            `Origin Point`, `Destination Point`
        )
    ) %>%
    collect()


df <- ds_new %>%
  group_by(Year, Month, `Day of Week`, `Hour of Day`, start_CBD, end_CBD) %>%
  summarize(
    hour_sum = sum(`Estimated Average Ridership`)
  ) %>%
  mutate(`Day of Week` = factor(`Day of Week`, levels = 
                                  c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%
  arrange(Year, Month, `Day of Week`, `Hour of Day`)

write.csv(df, paste0(base_path, "/data/processed/mta-hourly-cbd.csv"), row.names = FALSE)

df <- df %>%
  group_by(Year, Month, `Day of Week`, start_CBD, end_CBD) %>%
  summarize(
    daily_sum = sum(hour_sum)
  )
write.csv(df, file.path(base_path, "/data/processed/mta-daily-cbd.csv"), row.names = FALSE)
```